---
title: "1_Creating_time_series_objects.RMD"
author: "Russ Conte"
date: "11/26/2021"
output: html_document
---

We will begin by importing the required libraries into r.

```{r Import the data}
library(nycflights13)
library(tidyverse)
library(fpp3)
library(clock)

```

Our first step will be converting the nycflights13 data set into a tsibble, so time series forcasting can be done with the data. This will include setting up a date column to make the analysis much easier and faster

```{r Import the data, convert to a tsibble}

flights <- nycflights13::flights %>% 
    mutate(date = date_build(year = year, month = month, day = day))

flight1 <- flights %>% 
  mutate(date = date_build(year = year, month = month, day = day))

flight1 <- flight1 %>% 
  select(date, dep_time:distance)

```

# Clean up the data

The data has a few missing points, so it needs to be cleaned up before we can do full analysis with it. Specifically, it appears there are 8,255 rows with missing departure time, departure delay, arrival time, arrival delay and air time. The most likely explanation is that these flights were scheduled, but did not fly. Let's remove these rows from our data set.

```{r remove rows with flights listed but did not fly}
# 8,255 rows with missing departure times
flight1 %>%  filter(is.na(dep_time))

# remove rows with missing departure times
flight1 <- flight1 %>% 
  filter(!is.na(dep_time))

```

There are still 2,808 NAs in the data set. It appears all these flights did fly. Let's find these flights and see what we can do to clean everything up:

```{r Find the 2,808 rows with missing data but did fly}

sum(is.na(flight1)) # 2808

# Let's see where the NAs are located:
flight1 %>% 
  filter(!complete.cases(.))

# Just three columns contain all 2,808 NAs, so we'll build special data sets when we are making models to predict those values, otherwise we'll use the full data set when we create models to predict the values where we do have full data (such as predicting departure time):
sum(is.na(flight1$arr_time), is.na(flight1$arr_delay), is.na(flight1$air_time)) # 2,808

```

Just remember, when building models that include arrival time, arrival delay, and air time, to remove the rows with NAs first.

# Aggregate the data and create summary variables. The mean flight number column is used to create a key for the tsibble (next step)

```{r Aggregate the data}
flight1 <- flight1 %>% 
  group_by(date) %>% 
  summarise(
    flights = n(),
    mean_dep_time = mean(dep_time),
    mean_dep_delay = mean(dep_delay),
    mean_arrival_time = mean(arr_time, na.rm = TRUE),
    mean_arrival_delay = mean(arr_delay, na.rm = TRUE),
    mean_air_time = mean(air_time, na.rm = TRUE),
    mean_distance = mean(distance),
    key = 100
  ) %>% 
  ungroup()

```

# Calculate the most common carrier, destination and origin, and add those to the data set

```{r}
# Determine most common destination per day
mcd <- flights %>% 
  select(date, dest) %>% 
  group_by(date, dest) %>% 
  summarise(most_common_destination = n()) %>% 
  arrange(desc(most_common_destination)) %>% 
  filter(row_number() == 1) %>% 
  arrange(date)

flight1 <- flight1 %>% 
  mutate(most_common_destination = mcd$dest)

# Determine most common carrier per day
mcc <- flights %>% 
  select(date, carrier) %>% 
  group_by(date, carrier) %>% 
  summarise(most_common_carrier = n()) %>% 
  arrange(desc(most_common_carrier)) %>% 
  filter(row_number() == 1) %>% 
  arrange(date)

flight1 <- flight1 %>% 
  mutate(most_common_carrier = mcc$carrier)

# Determine most common origin per day
mco <- flights %>% 
  select(date, origin) %>% 
  group_by(date, origin) %>% 
  summarise(most_common_origin = n()) %>% 
  arrange(desc(most_common_origin)) %>% 
  filter(row_number() == 1) %>% 
  arrange(date)

flight1 <- flight1 %>% 
  mutate(most_common_origin = origin)


```


# Create a tsibble, from which we can do forecasting

```{r Convert flight1 into a tsibble}

flight1 <- flight1 %>% 
  as_tsibble(index = date, key = key) %>% 
  arrange(date) %>% 
  ungroup()

```

# Create a plot of the number of flights per day:

```{r plot of number of flights per day}
flight1 %>% 
  autoplot(flights)
```

# Create a plot for the mean departure time:

```{r plot mean departure time}

flight1 %>%
  autoplot(mean_dep_time)
```

# Superimpose mean departure time and mean departure delay:

```{r Combine mean departure time and mean departure delay}

flight1 %>% 
  autoplot(mean_dep_time) +
  geom_line(aes(y = mean_dep_delay))

```

# Exploring relationships <i>between</i> variables in our time series

```{r Seasonal plots of flights}

flight1 %>% 
 ggplot(aes(x = mean_dep_time, y = mean_dep_delay)) +
  geom_point() +
  labs(x = "Average departure time",
       y = "Average departure delay")

```

# Printing correlation between variables in our time series

```{r Pairs of correlations between all values in flight1}
flight1 %>% 
  GGally::ggpairs()
```

Discussion: Most correlations are <0.2, but a few are much larger:

The correlation between mean_distance and date = 0.473. This implies the later in the year a person flies, the farther they will travel.

The largest correlation is between mean of arrival delay, and mean of departure delay. That value is 0.944. There is a 94% probability in this data set that if a plane leaves late, it will arrive late.

# Lag plots

```{r Lagged scatterplots of flights}

flight1 %>% 
  gg_lag(flights, geom = "point") +
  labs(x = "lag(flights, k)")

```

(from the text) Each graph shows $y_t$ plotted against $y_{t-k}$ for different values of $k$

# Autocorrelation

(from the text) Just as correlation measures the extent of a linear relationship between two variables, autocorrelation measures the linear relationship between lagged values of a time series.

There are several autocorrelation coefficients, corresponding to each panel in the lag plot. For example, $r_1$ measures the relationship between $y_t$ and  $y_{t-1}$, $r_2$ measures the relationship between $y_t$ and $y_{t-2}$, and so on.

```{r Autocorrelation values for Flight 1 data set}
flight1 %>% ACF(flights, lag_max = 10)
```

We can view the values in the autocorrelation function:

```{r}
flight1 %>% 
  ACF(flights) %>% 
  autoplot() +
  labs(title = "Flights out of New York City in 2013, by day")
```

The strongest lags are at 7. 14, and 21 days. This is clearly due to the weekly nature of the business. $r_1$ is strong because the number of flights on any given day is strongly related to the number of flights the previous day.

(from the text) The dashed blue lines indicate whether the correlations are significantly different from zero.

# Trend and seasonality in our plot:

(from the text) When data have a trend, the autocorrelations for small lags tend to be large and positive because observations nearby in time are also nearby in value. So the ACF of a trended time series tends to have positive values that slowly decrease as the lags increase.

When data are seasonal, the autocorrelations will be larger for the seasonal lags (at multiples of the seasonal period) than for other lags.

If the values of $r_1$ through $r_t$ are all within the dotted blue lines, the results show white noise.